Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
carefully crafted page breaks on pages 1 to 3

\documentclass[nonatbib,blockstyle,times]{sigplanconf}
  % at 9pt anything but block style is unreadable

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And in~a Plan7 model, as discussed above, 
there are seven transitions into each node:
\smallverbatiminput{tprob-tprobs}
%\smallverbatiminput{tprobs}
In~our code, we select from this 7-tuple using \texttt{aScore}, a
function which satisfies
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$},
where $s$~and~$\hat s$ form one of the 7~pairs above.
Such a 7-tuple of scores is stored in each node in
the model.
A~node also contains tables of emission probabilities
$e_{M_j}$~and~$e_{I_j}$.
These tables are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

@@

% The bibliography should be embedded for final submission.

\bibliography{mrfy_experience_report}

>>>>

\documentclass[preprint,nonatbib,blockstyle,times]{sigplanconf}
  % at 9pt anything but block style is unreadable

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
\ifpagetuning
``Plan7'' implies that
\else
The model gets its name because
\fi
 there are exactly~7 possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And in~a Plan7 model, as discussed above, 
there are 7~transitions into each node,
each labeled $a_{s\hat s}$,
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
%\smallverbatiminput{tprobs}
To~choose a transition score from this 7-tuple, we use
function \texttt{aScore}, 
where
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
A~7-tuple is stored in each node in
the model.
Each node also contains tables of emission probabilities
$e_{M_j}$~and~$e_{I_j}$, which are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

@@

\iffalse
\bibliography{mrfy_experience_report}
\else
\begin{thebibliography}{9}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Cowen et~al.(2002)Cowen, Bradley, Menke, King, and
  Berger]{Cowen:2002p588}
Lenore Cowen, Philip Bradley, Matt Menke, Jonathan King, and Bonnie Berger.
\newblock {Predicting the beta-helix fold from protein sequence data}.
\newblock \emph{Journal of Computational Biology}, 2002.

\bibitem[Daniels et~al.(2012)Daniels, Hosur, Berger, and Cowen]{Daniels:2012}
Noah Daniels, Raghavendra Hosur, Bonnie Berger, and Lenore Cowen.
\newblock {SMURFLite: combining simplified Markov random fields with simulated
  evolution improves remote homology detection for beta-structural proteins
  into the twilight zone}.
\newblock \emph{Bioinformatics (to appear)}, 2012.

\bibitem[Durbin et~al.(1998)Durbin, Eddy, Krogh, and Mitchison]{Durbin:1998wz}
Rirchard Durbin, Sean Eddy, Anders Krogh, and Graeme Mitchison.
\newblock \emph{{Biological Sequence Analysis: Probabilistic Models of Proteins
  and Nucleic Acids}}.
\newblock Cambridge University Press, May 1998.

\bibitem[Eddy(1998)]{Eddy:1998ut}
Sean Eddy.
\newblock {Profile hidden Markov models}.
\newblock \emph{Bioinformatics}, 14:\penalty0 755--763, 1998.

\bibitem[Hughes(1989)]{hughes:why}
John Hughes.
\newblock Why functional programming matters.
\newblock \emph{The Computer Journal}, 32\penalty0 (2):\penalty0 98--107, April
  1989.

\bibitem[Menke et~al.(2008)Menke, Berger, and Cowen]{Menke:2008wu}
Matthew Menke, Bonnie Berger, and Lenore Cowen.
\newblock {Matt: local flexibility aids protein multiple structure alignment}.
\newblock \emph{PLoS Computational Biology}, 2008.

\bibitem[Menke et~al.(2010)Menke, Berger, and Cowen]{Menke:2010ti}
Matthew Menke, Bonnie Berger, and Lenore Cowen.
\newblock {Markov random fields reveal an N-terminal double beta-propeller
  motif as part of a bacterial hybrid two-component sensor system}.
\newblock \emph{Proceedings of the National Academy of Science}, 2010.

\bibitem[Sansom and Peyton~Jones(1997)]{sansom-pj}
Patrick Sansom and Simon~L Peyton~Jones.
\newblock Formally based profiling for higher-order functional languages.
\newblock \emph{ACM TOPLAS}, 19\penalty0 (2):\penalty0 334--385, 1997.

\bibitem[Viterbi(1967)]{Viterbi:1967hq}
Andrew Viterbi.
\newblock {Error bounds for convolutional codes and an asymptotically optimum
  decoding algorithm}.
\newblock \emph{IEEE Transactions on Information Theory}, 13\penalty0
  (2):\penalty0 260--269, April 1967.

\end{thebibliography}
\fi
